A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-07 paper/paper.tex
full draft of semantics section

\begin{construction}[Soundness and Completeness]
  For any choice of connectives, the syntax of parameterized \ohol{}
  defines a hyperdoctrine of VDCrs that is an initial object in the
  category of hyperdoctrines of VDCrs with chosen universal
  constructions and functors preserving those universal constructions.
\end{construction}
\begin{proof}
  Most of the details of the proof are straightforward. In the
  appendix we demonstrate the syntactic elements of the proof which
  are that the syntax itself is a hyperdoctrine of VDCrs, i.e., that
  the notions of substitution are appropriately functorial.
\end{proof}

@@

\begin{construction}[(\citet{shulman-enriched-indexed})]
  Given any functor $\mathcal V : S^o \to \SymMonCat$ such that $S$
  and $\mathcal V$ have sufficiently well-behaved (indexed)
  $\kappa$-products, then there is a virtual equipment $\mathcal
  V-\textrm{Cat}$ whose objects are locally $\kappa$-small $\mathcal
  V$-categories, small objects are $\kappa$-small $\mathcal
  V$-categories etc. This virtual equipment has been shown in
  loc. cit. to have objects satisfying all of the universal properties
  used in models of VETT except the graph of a profunctor.
\end{construction}

>>>>

An HVE is then precisely the structure corresponding to the
\emph{judgments} and actions of substitution in VETT.
\begin{construction}[Syntactic Model]
  \label{cons:syn-model}
  The syntax of VETT with any choice of which subset of connectives
  are included presents an HVE.
\end{construction}
\begin{proof}
  Straightforward, defining the category of families using the
  dependent type structure and the virtual equipment structure having
  ($\alpha$-equivalence classes of) syntactic categories as objects,
  functors/sets as vertical/horizontal arrows and interpreting
  compositions/restrictions as substitutions.  The biggest gap is in
  the definition of the 2-cells.

  A 2-cell from $(\alpha_1:\cat C_1;\alpha_2:\cat C_2\vdash
  R_1),(\alpha_2:\cat C_2;\alpha_3:\cat C_3),\ldots$ to $(\beta_1:\cat
  D_1;\beta_2\vdash S)$ with frames $\alpha_1:\cat C_1\vdash b_1 :
  \cat D_1$ and $\alpha_n:\cat C_n\vdash b_2 : \cat D_2$ given by a
  term $x_1:R_1,x_2:R_2\ldots \vdash s :
  S[b_1/\beta_1;b_2/\beta_2]$. Composition is then defined by
  substitution.
\end{proof}

Then the \emph{connectives} of VETT each precisely correspond to a
universal construction of structures in an HVE. The
$\Pi,\Sigma,\text{Id}$ types correspond to their standard semantics in
a CwF, products of categories interpreted as products in the vertical
category, and products of sets as products in the category of
pro-arrows and 2-cells. The units, tensor and covariant and
contravariant homs are modeled by the universal properties of the same
names, as described for instance in \citet{shulmanFrBicats}. The graph
of a profunctor is modeled by \emph{tabulators}
\citet{grandis-pare}. Finally, the covariant and contravariant
presheaf categories can be described as a weakening of the definition
of a Yoneda equipment from \citet{yosegi-boxes} to virtual double
categories. More detailed descriptions of these universal properties
are included in the appendix.
%
Then we can package up the soundness and completeness of this notion
of categorical model in the following modular initiality theorem.

\begin{theorem}[Initiality]
  The syntax of VETT with almost any choice of connectives presents an
  HVE that is initial in the category of HVEs with the chosen
  instances of the universal properties and functors that preserve
  such chosen instances. The only exception is that if units are
  included, then contravariant and covariant hom sets must be included
  as well.
\end{theorem}
\begin{proof}
  It is straightforward to extend the construction
  \ref{cons:syn-model} for any connective modularly, with the
  exception that the unit relies on the presence of hom sets in order
  to satisfy the ``distributivity'' requirement that its elimination
  can occur in any context. Then we can construct the unique morphism
  to any HVE by straightforward induction on syntax.
\end{proof}

@@

\begin{construction}[(\citet{shulman-enriched-indexed})]
  Given any functor $\mathcal V : S^o \to \SymMonCat$ such that $S$
  and $\mathcal V$ have sufficiently well-behaved (indexed)
  $\kappa$-products, then there is a virtual equipment $\mathcal
  V-\textrm{Cat}$ whose objects are locally $\kappa$-small $\mathcal
  V$-categories, small objects are $\kappa$-small $\mathcal
  V$-categories etc. This virtual equipment has objects satisfying all
  of the universal properties needed for a model of VETT.
\end{construction}
